---
title: "p8105_hw1_ml4420"
author: Mengjia Lyu
date:   2019-09-13
output: github_document
---

## Data Science Homework 1


```{r problem 1}
library(formattable)
library(tidyverse) #loading libraries

first_df = tibble(   #first data frame
  random_sample = rnorm(n = 8),
  vec_logical   = random_sample > 0,
  vec_char      = c("this", "is", "a", "character", "vector", "of", "length", "eight"),
  vec_factor    = factor(c("small", "medium", "large", "small", "medium", "large", "small", "medium")),
 )
```

Now let us try to take the mean of each variable in the data frame.

```{r problem 1 mean}


  mean(pull(first_df, random_sample)) #mean of random sample
  mean(pull(first_df, vec_logical)) #mean of logical variable
  mean(pull(first_df, vec_char))  #mean of char variable
  mean(pull(first_df, vec_factor)) #mean of factor variable


```
As can be seen from the results, taking the mean of the random sample or the logical vector works, but taking the mean of the character vector or the factor vector doesn't work.

Now let us try explicitly convert variables from one type to another.

```{r ECHO = T, warning = FALSE, results = 'hide'}

  as.numeric(pull(first_df, vec_logical)) #convert to numeric
  as.numeric(pull(first_df, vec_char))    
  as.numeric(pull(first_df, vec_factor))  
```
Converting the logical variable to numeric type, we get 0s and 1s. The reason is that logical variables only two type of values: TRUE or FALSE. 0s and 1s are enough to represent the values.

Converting the character variable to numeric type does not work and we get NAs which stand for missing values. The reason is that character vector stores character values whose encodings are fundamentally different from those of numeric values.

Converting the factor variable to numeric type, we get 1s, 2s and 3. The reason is that factor variables give numeric encodings to non-numeric variables and represent them by countable levels which can be rendered by numbers. The factor variable we created contain three dinstinct levels, which is why we get 1s, 2s and 3s.

This helps explain what happened when I tried to take the mean. If a variable cannot be or has not been converted into numeric type, it cannot have a mean as mean in essence is a numeric value. 

Since a character vector cannot be converted to numeric type, taking its mean does not make sense. 

Since the factor variable was not converted before taking its mean, missing values were returned. 

However, logical values are always binary which means they can always be represented by 0s and 1s. Therefore, its mean can be calculated numerically without conversions.


```{r problem_1 conversion and multiplication}

 result_1 = as.numeric(pull(first_df, vec_logical))   #convert to numeric
 product_1 = pull(first_df, random_sample) * result_1
 
 result_2 = as.factor(pull(first_df, vec_logical))    #convert to factor
 product_2 = pull(first_df, random_sample) * result_2
 
 result_3 = as.numeric(as.factor(pull(first_df, vec_logical)))  #convert to numeric
 product_3 = pull(first_df, random_sample) * result_3


```



## Problem 2

```{r problem 2}
second_df = tibble(   #second data frame
  x = rnorm(500),
  y = rnorm(500),
  vec_logical = x + y > 1,
  vec_numeric = as.numeric(vec_logical),
  vec_factor  = as.factor(vec_logical)
)
```

**Here is a short description of my vectors.** 
The dataset has **`r nrow(second_df)`** rows and **`r ncol(second_df)`** columns. 
The mean, median, and standard deviation of x is **`r mean(pull(second_df, x))`**, **`r median(pull(second_df, x))`** and **`r sd(pull(second_df, x))`** respectively.
The proportion of cases for which x + y > 1 is **`r percent(sum(pull(second_df, vec_logical))/500)`**

```{r scatterplot}
ggplot(second_df, aes(x = x, y = y, color = vec_logical)) + geom_point()
ggsave("scatterplot.pdf")  #export the first scatterplot to my project directory
ggplot(second_df, aes(x = x, y = y, color = vec_numeric)) + geom_point()
ggplot(second_df, aes(x = x, y = y, color = vec_factor)) + geom_point()
```
 
 
 **Comments:**
 When using the logical vector to color the plot, the color scale is either true or false. There are only two colors in the plot.
 When using the numeric vector to color the plot, the color scale is more continuous and there are a spectrum of colors in the plot.
 When using the factor vector to color the plot, the color scale is also either true or false. There are only two colors in the plot.
 

 

